---
title: "Multiple linear regression (MLR)"
author: "Prof. Maria Tackett"
date: "2025-01-30"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 210 - Spring 2025](https://sta210-sp25.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 07-mlr-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

## Announcements {.midi}

-   Lab 01 due **TODAY** at 11:59pm
-   Team labs start on Monday
-   [Click here](https://prodduke.sharepoint.com/:p:/s/ARCStaff839/EZ4PKTRTlCVMpFZiR6XoRycB4UUlRMuI2_Rda9hKxNZtsA) to learn more about the Academic Resource Center
-   [Statistics experience](../hw/stats-experience.html) due Tuesday, April 15

## Computational setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(patchwork)
library(knitr)
library(kableExtra)
library(colorblindr)
library(palmerpenguins)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

# Considering multiple variables

## Data: Palmer penguins {.midi}

The data set contains data for penguins found on three islands in the Palmer Archipelago, Antarctica. Data were collected and made available by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station, Antarctica LTER](https://pal.lternet.edu/), a member of the [Long Term Ecological Research Network](https://lternet.edu/). These data can be found in the **palmerpenguins** R package.

<br>

. . .

```{r}
#| echo: false

penguins <- penguins |>
  select(body_mass_g, flipper_length_mm, bill_length_mm, species) |>
  drop_na()

penguins
```

## Variables

**Predictors**:

-   `bill_length_mm`: Bill length in millimeters
-   `flipper_length_mm`: Flipper length in millimeters
-   `species`: Adelie, Gentoo, or Chinstrap species

**Outcome**: `body_mass_g`: Body mass in grams

<br>

**The goal of this analysis is to use the bill length, flipper length, and species to predict body mass.**

## Outcome: `body_mass_g`

```{r}
#| echo: false
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density(fill = "steelblue") +
  labs(x = "Body mass in grams",
    title = "Distribution of body mass of Palmer Penguins")
```

```{r}
#| echo: false
penguins |>
  summarise(
    min = min(body_mass_g),
    median = median(body_mass_g),
    max = max(body_mass_g),
    iqr = IQR(body_mass_g)
  ) |>
  kable()
```

## Predictors {.small}

```{r}
#| echo: false
#| out.width: "100%"
p1 <- ggplot(penguins, aes(y = species)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Species", 
       y = "")

p2 <- ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(fill = "steelblue", color = "black") +
  labs(title = "",
       x = "Flipper length in mm")

p3 <- ggplot(penguins, aes(x = bill_length_mm)) +
  geom_histogram(fill = "steelblue", color = "black") +
  labs(title = "",
       x = "Bill length in mm")

p1 + p2 / p3
```

## Outcome vs. predictors {.small}

```{r}
#| echo: false
p4 <- ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    y = "Body mass in grams",
    x = "Species"
  )

p5 <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(color = "steelblue") +
  labs(
    y = NULL,
    x = "Flipper length in mm"
  )


p6 <- ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(color = "steelblue") +
  labs(
    y = NULL,
    x = "Bill length in mm"
  )

p4 + p5 / p6
```

. . .

::: question
Why do we want to use a single model with all the predictors instead of 3 separate models?
:::

## Single vs. multiple predictors

So far we've used a *single predictor variable* to understand variation in a quantitative response variable

. . .

Now we want to use *multiple predictor variables* to understand variation in a quantitative response variable

# Multiple linear regression

## Multiple linear regression (MLR)

Based on the analysis goals, we will use a **multiple linear regression** model of the following form

$$
\begin{aligned}\widehat{\text{body_mass_g}} ~ = 
\hat{\beta}_0 & + \hat{\beta}_1 \times \text{flipper_length_mm} \\ & + \hat{\beta}_2 \times  \text{species}_1  \\
&+\hat{\beta}_3 \times \text{species}_2  \\ &+ \hat{\beta}_4 \times \text{bill_length_mm}
\end{aligned}
$$

Similar to simple linear regression, this model assumes that at each combination of the predictor variables, the values `body_mass_g` follow a Normal distribution.

## Multiple linear regression

**Recall:** The simple linear regression model assumes

$$
Y|X\sim N(\beta_0 + \beta_1 X, \sigma_{\epsilon}^2)
$$

. . .

**Similarly:** The multiple linear regression model assumes

$$
Y|X_1, X_2, \ldots, X_p \sim N(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p, \sigma_{\epsilon}^2)
$$

## Multiple linear regression

At any combination of the predictors, the mean value of the response $Y$, is

$$
\mu_{Y|X_1, \ldots, X_p} = \beta_0 + \beta_1 X_{1} + \beta_2 X_2 + \dots + \beta_p X_p
$$

. . .

Using multiple linear regression, we can estimate the mean response for any combination of predictors

$$
\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X_{1} + \hat{\beta}_2 X_2 + \dots + \hat{\beta}_p X_{p}
$$

## Model fit

```{r}
#| echo: true

penguin_fit <- lm(body_mass_g ~ flipper_length_mm + species + 
                bill_length_mm, data = penguins)

tidy(penguin_fit) |>
  kable(digits = 3)
```

## Model equation

$$
\begin{align}\widehat{\text{body_mass_g}} =  -3904.387 &+27.429 \times \text{flipper_length_mm}\\
& -748.562 \times \text{Chinstrap}\\  
&+ 90.435 \times \text{Gentoo}\\
&+ 61.736 \times \text{bill_length_mm}
\end{align}
$$

::: callout-note
We will talk about why there are two terms in the model for `species` shortly!
:::

## Interpreting $\hat{\beta}_j$

-   The estimated coefficient $\hat{\beta}_j$ is the expected change in the mean of $Y$ when $X_j$ increases by one unit, <i>holding the values of all other predictor variables constant</i>.

. . .

-   **Example:** The estimated coefficient for `flipper_length_mm` is 27.429. This means for each additional millimeter in a penguin's flipper length, its body mass is expected to be greater by 27.429 grams, on average, holding species and bill length constant.

## Prediction

::: question
What is the predicted body mass for a Gentoo penguin with a flipper length of 200 millimeters and bill length of 45 millimeters?
:::

<br>

```{r}
#| echo: true

-3904.387 + 27.429 * 200 - 748.562 * 0 + 90.435 * 1 + 61.736 * 45

```

<br>

. . .

The predicted body mass for a Gentoo penguin with a flipper length of 200 millimeters and bill length of 45 millimeters is `r round(-3904.387 + 27.429 * 200 - 748.562 * 0 + 90.435 * 1 + 61.736 * 45, 3)` grams.

## Prediction, revisited {.midi}

Just like with simple linear regression, we can use the `predict()` function in R to calculate the appropriate intervals for our predicted values:

```{r}
#| echo: true

new_penguin <- tibble(
  flipper_length_mm  = 200, 
  species = "Gentoo", 
  bill_length_mm = 45
)

predict(penguin_fit, new_penguin)
```

::: callout-note
Difference in predicted value due to rounding the coefficients on the previous slide.
:::

## Confidence interval for $\hat{\mu}_y$

::: question
Calculate a 90% confidence interval for the **estimated mean body mass** a Gentoo penguins with a flipper length of 200 millimeters and bill length of 45 millimeters.
:::

<br>

```{r}
#| echo: true

predict(penguin_fit, new_penguin, interval = "confidence", 
        level = 0.90) |> 
  kable(digits = 3)
```

## Prediction interval for $\hat{y}$

::: question
Calculate a 90% confidence interval for the **estimated body mass** for an individual Gentoo penguin with a flipper length of 200 millimeters and bill length of 45 millimeters.
:::

<br>

```{r}
#| echo: true

predict(penguin_fit, new_penguin, interval = "prediction", 
        level = 0.90) |>
  kable(digits = 3)
```

## Cautions

-   **Do not extrapolate!** Because there are multiple predictor variables, there is the potential to extrapolate in many directions
-   The multiple regression model only shows **association, not causality**
    -   To show causality, you must have a carefully designed experiment or carefully account for confounding variables in an observational study

# Types of predictors

# Categorical predictors

## Indicator variables

-   Suppose there is a categorical variable with $k$ categories (levels)

-   We can make $k$ indicator variables - one indicator for each category

-   An **indicator variable** takes values 1 or 0

    -   1 if the observation belongs to that category
    -   0 if the observation does not belong to that category

## Indicator variables for `species`

```{r}
#| echo: true

penguins <- penguins |>
  mutate(
    adelie = if_else(species == "Adelie", 1, 0),
    chinstrap = if_else(species == "Chinstrap", 1, 0),
    gentoo = if_else(species == "Gentoo", 1, 0)
  )
```

. . .

```{r}
#| echo: false
penguins |>
  select(species, adelie, chinstrap, gentoo) |>
  slice(1, 152, 283)
```

## Indicators in the model {.midi}

-   We will use $k-1$ of the indicator variables in the model.
-   The **baseline** is the category that doesn't have a term in the model.
-   The coefficients of the indicator variables in the model are interpreted as the expected change in the response compared to the baseline, holding all other variables constant.
-   This approach is also called **dummy coding**.

. . .

```{r}
penguins |>
  select(species, chinstrap, gentoo) |>
  slice(1, 152, 283)
```

## Interpreting `species` {.midi}

```{r}
#| echo: false
tidy(penguin_fit, conf.int  = T) |>
  kable(digits = 3) |>
  row_spec(c(3,4), background = "#dce5b2")
```

. . .

::: incremental
-   The baseline category is `Adelie`.
-   Penguins from the Chinstrap species are expected to have body mass that is 748.562 grams less on average, compared to penguins from the Adelie species, holding flipper length and bill length constant.
:::

. . .

::: question
Interpret the coefficient of `Gentoo` in the context of the data.
:::

# Transforming quantitative predictors

## Interpreting the intercept {.midi}

```{r}
#| echo: false
tidy(penguin_fit, conf.int = TRUE) |>
  kable(digits = 3)
```

. . .

<br>

::: question
The intercept -3904.387 is not meaningful. Let's transform some variables to make this intercept meaningful.
:::

## Centering {.midi}

-   Centering a quantitative predictor means shifting every value by some constant $C$

$$
X_{cent} = X  - C
$$

-   One common type of centering is **mean-centering**, in which every value of a predictor is shifted by its mean

-   Only quantitative predictors are centered

-   Center all quantitative predictors in the model for ease of interpretation

::: question
What is one reason one might want to center the quantitative predictors? What is are the units of centered variables?
:::

## Centering {.midi}

Use the `scale()` function with `center = TRUE` and `scale = FALSE` to mean-center variables

```{r}
#| output-location: fragment 

penguins <- penguins |>
  mutate(flipper_length_cent = scale(flipper_length_mm, center = TRUE, scale = FALSE), 
         bill_length_cent = scale(bill_length_mm, center = TRUE, scale = FALSE))
```

## Original vs. mean-centered variable

::::: columns
::: {.column width="50%"}
**Original variable**

```{r}
#| echo: false 
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_density(fill = "steelblue") +
  labs(x = NULL, title = "Flipper length in mm")
```

<br>

```{r}
#| echo: false
penguins |>
  summarise(Mean = mean(flipper_length_mm),
            SD = sd(flipper_length_mm)) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
**Mean-centered variable**

```{r}
#| echo: false 
ggplot(penguins, aes(x = flipper_length_cent)) +
  geom_density(fill = "darkcyan") +
  labs(x = NULL, title = "Mean-centered Flipper length in mm")
```

<br>

```{r}
#| echo: false 

penguins |>
  summarise(Mean = mean(flipper_length_cent),
            SD = sd(flipper_length_cent)) |>
  kable(digits = 3)
```
:::
:::::

<br>

. . .

## Using mean-centered variables in the model

::: question
How do you expect the model to change if we use `flipper_length_cent` and `bill_length_cent` in the model?
:::

. . .

```{r}
#| echo: false
penguin_cent_fit <- lm(body_mass_g ~ flipper_length_cent + species +
                    bill_length_cent, data = penguins)

tidy(penguin_cent_fit, conf.int = T) |>
  kable(digits = 3)
```

## Original vs. mean-centered model {.small}

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
tidy(penguin_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
tidy(penguin_cent_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::
:::::

::: question
What has changed? What is the same?
:::

## Standardizing {.midi}

-   Standardizing a quantitative predictor mean shifting every value by the mean and dividing by the standard deviation of that variable

$$
X_{std} = \frac{X - \bar{X}}{S_X}
$$

-   Only quantitative predictors are standardized

-   Standardize all quantitative predictors in the model for ease of interpretation

::: question
What is one reason one might want to standardize the quantitative predictors? What is are the units of standardized variables?
:::

## Standardizing

Use the `scale()` function with `center = TRUE` and `scale = TRUE` to standardized variables

<br>

```{r}
#| output-location: fragment 
#| 
penguins <- penguins |>
  mutate(flipper_length_std = scale(flipper_length_mm, center = TRUE, scale = TRUE), 
         bill_length_std = scale(bill_length_mm, center = TRUE, scale = TRUE))
```

## Original vs. standardized variable

::::: columns
::: {.column width="50%"}
**Original variable**

```{r}
#| echo: false 
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_density(fill = "steelblue") +
  labs(x = NULL, title = "Flipper length in mm")
```

<br>

```{r}
#| echo: false

penguins |>
  summarise(Mean = mean(flipper_length_mm),
            SD = sd(flipper_length_mm)) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
**Standardized variable**

```{r}
#| echo: false 
ggplot(penguins, aes(x = flipper_length_std)) +
  geom_density(fill = "darkcyan") +
  labs(x = NULL, title = "Standardized Flipper length in mm")
```

<br>

```{r}
#| echo: false

penguins |>
  summarise(Mean = mean(flipper_length_std),
            SD = sd(flipper_length_std)) |>
  kable(digits = 3)
```
:::
:::::

<br>

. . .

## Using standardized variables in the model

::: question
How do you expect the model to change if we use `flipper_length_std` and `bill_length_std` in the model?
:::

. . .

```{r}
#| echo: false
penguin_std_fit <- lm(body_mass_g ~ flipper_length_std + species +
                    bill_length_std, data = penguins)

tidy(penguin_std_fit, conf.int = T) |>
  kable(digits = 3)
```

## Original vs. standardized model

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
tidy(penguin_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
tidy(penguin_std_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::
:::::

::: question
What has changed? What is the same?
:::

# Interaction terms

## Interaction terms

-   Sometimes the relationship between a predictor variable and the response depends on the value of another predictor variable.
-   This is an **interaction effect**.
-   To account for this, we can include **interaction terms** in the model.

## Bill length versus species

If the lines are not parallel, there is indication of a potential **interaction effect**, i.e., the slope of bill length may differ based on the species.

```{r}
#| echo: false

p1 <- ggplot(penguins, 
             aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Bill length in millimeters",
    y = "Body mass in grams"
  )

p2 <- ggplot(penguins, 
             aes(x = bill_length_mm, y = body_mass_g,
                 color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Bill length in millimeters", y = NULL, color = NULL) +
  theme(legend.position = c(0.3, 0.9)) +
  scale_color_OkabeIto()

p1 + p2 +
  plot_annotation(title = "Body mass versus bill length")
```

## Interaction term in model {.smaller}

```{r}
#| echo: true
penguin_int_fit <- lm(body_mass_g ~ flipper_length_mm + species + bill_length_mm + species * bill_length_mm,
      data = penguins)
```

```{r}
#| echo: false
tidy(penguin_int_fit) |>
  kable(digits = 3) |>
  row_spec(c(6,7), background = "#dce5b2")
```

## Interpreting interaction terms

-   What the interaction means: The effect of bill length on the body mass is 41.035 less when the penguin is from the Chinstrap species compared to the effect for the Adelie species, holding all else constant.
-   Interpreting `bill_length_mm` for Chinstrap: For each additional millimeter in bill length, we expect the body mass of Chinstrap penguins to increase by 31.657 grams (72.692 - 41.035), holding all else constant.

## Summary

<br>

::: question
In general, how do

indicators for categorical predictors impact the model equation?

interaction terms impact the model equation?
:::

## Recap

-   Introduced multiple linear regression

-   Interpreted coefficients in the multiple linear regression model

-   Calculated predictions and associated intervals for multiple linear regression models

-   Mean-centered and standardized quantitative predictors

-   Used indicator variables for categorical predictors

-   Used interaction terms
