---
title: "MLR: Model comparison + Inference"
author: "Prof. Maria Tackett"
footer: "[🔗 STA 210 - Spring 2025](https://sta210-sp25.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 10-mlr-inference-conditions-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

## Announcements

-   HW 02 due TODAY at 11:59pm

-   Lab 03 due Thursday, February 13 at 11:59pm

-   Project topics due Sunday, February 16 at 11:59pm

-   [Statistics experience](../hw/stats-experience.html) due Tuesday, April 15

## Exam 01

-   50 points total
    -   in-class: 35-40 points

    -   take-home: 10 - 15 points
-   In-class (35 -40 pts): 75 minutes during February 18 lecture
-   Take-home (10 -15 pts): released after class on Tuesday
-   If you miss any part of the exam for an excused absence (with academic dean’s note or other official documentation), your Exam 02 score will be counted twice

## Tips for studying

-   Review exercises in AEs and assignments, asking “why” as you review your process and reasoning

    -   e.g., Why do we include “holding all else constant” in interpretations?

-   Focus on understanding not memorization

-   Explain concepts / process to others

-   Ask questions in office hours

-   Review lecture recordings as needed

## Topics

::: nonincremental
-   Model comparison AE
-   Inference for multiple linear regression
:::

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(patchwork)
library(knitr)
library(kableExtra)
library(countdown)
library(rms)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Model comparison

## RMSE

$$
RMSE = \sqrt{\frac{\sum_{i=1}^n(y_i - \hat{y}_i)^2}{n}} = \sqrt{\frac{\sum_{i=1}^ne_i^2}{n}}
$$

::: incremental
-   Ranges between 0 (perfect predictor) and infinity (terrible predictor)

-   Same units as the response variable

-   The value of RMSE is more useful for comparing across models than evaluating a single model
:::

## $R^2$ and Adjusted $R^2$

$$R^2 = \frac{SSM}{SST} = 1 - \frac{SSR}{SST}$$

<br>

. . .

$$Adj. R^2 = 1 - \frac{SSR/(n-p-1)}{SST/(n-1)}$$

where

-   $n$ is the number of observations used to fit the model

-   $p$ is the number of terms (not including the intercept) in the model

# Application exercise

::: appex
📋 [sta210-sp25.netlify.app/ae/ae-06-model-compare](../ae/ae-06-model-compare.html)
:::

<br>

[Click here](https://docs.google.com/presentation/d/16rXxcbXq2ycI7LpxMNmyclBwkdxbhQY_GS4GYX316bE/edit?usp=sharing) to share your group's response.

# Inference for multiple linear regression

## Modeling workflow

-   Split data into training and test sets.

-   Fit, evaluate, and compare candidate models. Choose a final model based on summary of cross validation results.

-   Refit the model using the entire training set and do "final" evaluation on the test set (make sure you have not overfit the model).

    -   Adjust as needed if there is evidence of overfit.

-   Use model fit on training set for inference and prediction.

## Data: `rail_trail` {.midi}

::: nonincremental
-   The Pioneer Valley Planning Commission (PVPC) collected data for ninety days from April 5, 2005 to November 15, 2005.
-   Data collectors set up a laser sensor, with breaks in the laser beam recording when a rail-trail user passed the data collection station.
:::

```{r}
#| echo: false
rail_trail <- read_csv(here::here("slides", "data/rail_trail.csv"))
rail_trail |> slice(1:8)
```

Source: [Pioneer Valley Planning Commission](http://www.fvgreenway.org/pdfs/Northampton-Bikepath-Volume-Counts%20_05_LTA.pdf) via the **mosaicData** package.

## Variables {.midi}

**Response**:

`volume` estimated number of trail users that day (number of breaks recorded)

. . .

**Predictors**

::: nonincremental
-   `hightemp` daily high temperature (in degrees Fahrenheit)
-   `avgtemp` average of daily low and daily high temperature (in degrees Fahrenheit)
-   `season` one of "Fall", "Spring", or "Summer"
-   `cloudcover` measure of cloud cover (in oktas)
-   `precip` measure of precipitation (in inches)
-   `day_type` one of "weekday" or "weekend"
:::

# Conduct a hypothesis test for $\beta_j$

## Review: Simple linear regression (SLR)

```{r}
#| echo: false
ggplot(rail_trail, aes(x = hightemp, y = volume)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "High temp (F)", y = "Number of riders")
```

## SLR model summary

```{r}
rt_slr_fit <- lm(volume ~ hightemp, data = rail_trail)

tidy(rt_slr_fit) |> kable(digits = 2)
```

## SLR hypothesis test {.midi}

```{r}
#| echo: false

tidy(rt_slr_fit) |> kable(digits = 2)
```

1.  **Set hypotheses:** $H_0: \beta_1 = 0$ vs. $H_a: \beta_1 \ne 0$

. . .

2.  **Calculate test statistic and p-value:** The test statistic is $t= 6.72$ . The p-value is calculated using a $t$ distribution with 88 degrees of freedom. The p-value is $\approx 0$ .

. . .

3.  **State the conclusion:** The p-value is small, so we reject $H_0$. The data provide strong evidence that high temperature is a helpful predictor for the number of daily riders, i.e. there is a linear relationship between high temperature and number of daily riders.

## Multiple linear regression

```{r}
rt_mlr_main_fit <- lm(volume ~ hightemp + season, data = rail_trail)

tidy(rt_mlr_main_fit) |> kable(digits = 2)
```

## Multiple linear regression

The multiple linear regression model assumes $$Y|X_1, X_2,  \ldots, X_p \sim N(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p, \sigma_\epsilon^2)$$

<br>

. . .

For a given observation $(x_{i1}, x_{i2}, \ldots, x_{ip}, y_i)$, we can rewrite the previous statement as

$$y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} + \epsilon_{i} \hspace{10mm} \epsilon_i \sim N(0,\sigma_{\epsilon}^2)$$

------------------------------------------------------------------------

## Estimating $\sigma_\epsilon$ {.midi}

For a given observation $(x_{i1}, x_{i2}, \ldots,x_{ip}, y_i)$ the residual is $$e_i = y_{i} - (\hat{\beta}_0 + \hat{\beta}_1 x_{i1} + \hat{\beta}_{2} x_{i2} + \dots + \hat{\beta}_p x_{ip})$$

<br>

. . .

The estimated value of the regression standard error , $\sigma_{\epsilon}$, is

$$\hat{\sigma}_\epsilon  = \sqrt{\frac{\sum_{i=1}^ne_i^2}{n-p-1}}$$

. . .

As with SLR, we use $\hat{\sigma}_{\epsilon}$ to calculate $SE(\hat{\beta}_j)$, the standard error of the coefficient for predictor $x_j$. See [Matrix Form of Linear Regression](https://github.com/STA210-Sp19/supplemental-notes/blob/master/regression-basics-matrix.pdf) for more detail.

## MLR hypothesis test: hightemp {.midi}

1.  **Set hypotheses:** $H_0: \beta_{hightemp} = 0$ vs. $H_a: \beta_{hightemp} \ne 0$, given `season` is in the model

. . .

2.  **Calculate test statistic and p-value:** The test statistic is $t = 6.43$. The p-value is calculated using a $t$ distribution with 86 $(n - p - 1)$ degrees of freedom. The p-value is $\approx 0$.

. . .

3.  **State the conclusion:** The p-value is small, so we reject $H_0$. The data provide strong evidence that high temperature for the day is a useful predictor in a model that already contains the season as a predictor for number of daily riders.

## Interaction terms {.midi}

```{r}
#| echo: false

rt_mlr_int_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(volume ~ hightemp + season + hightemp * season, data = rail_trail)

tidy(rt_mlr_int_fit) |> kable(digits = 2)

```

<br>

::: question
Do the data provide evidence of a significant interaction effect? Comment on the significance of the interaction terms.
:::

# Confidence interval for $\beta_j$

## Confidence interval for $\beta_j$ {.midi}

::: incremental
-   The $C\%$ confidence interval for $\beta_j$ $$\hat{\beta}_j \pm t^* SE(\hat{\beta}_j)$$ where $t^*$ follows a $t$ distribution with $n - p - 1$ degrees of freedom.
-   **Generically**: We are $C\%$ confident that the interval LB to UB contains the population coefficient of $x_j$.
-   **In context:** We are $C\%$ confident that for every one unit increase in $x_j$, $y$ changes by LB to UB units, on average, holding all else constant.
:::

## Confidence interval for $\beta_j$

```{r}
tidy(rt_mlr_main_fit, conf.int = TRUE, conf.level = 0.95) |>
  kable(digits= 2)
```

## CI for `hightemp` {.midi}

```{r}
#| echo: false

tidy(rt_mlr_main_fit, conf.int = TRUE) |>
  kable(digits = 2)
```

<br>

We are 95% confident that for every degree Fahrenheit the day is warmer, the number of riders increases by 5.21 to 9.87, on average, holding season constant.

## CI for `seasonSpring` {.midi}

```{r}
#| echo: false

tidy(rt_mlr_main_fit, conf.int = TRUE) |>
  kable(digits = 2)
```

<br>

We are 95% confident that the number of riders on a Spring day is lower by 63.1 to higher by 73.4 compared to a Fall day, on average, holding high temperature for the day constant.

. . .

::: question
Is `season` a significant predictor of the number of riders, after accounting for high temperature?
:::

# Inference pitfalls

## Large sample sizes

<br>

:::: callout-caution
If the sample size is large enough, the test will likely result in rejecting $H_0: \beta_j = 0$ even $x_j$ has a very small effect on $y$.

::: nonincremental
-   Consider the **practical significance** of the result not just the statistical significance.

-   Use the confidence interval to draw conclusions instead of relying only p-values.
:::
::::

## Small sample sizes

<br>

:::: callout-caution
If the sample size is small, there may not be enough evidence to reject $H_0: \beta_j=0$.

::: nonincremental
-   When you fail to reject the null hypothesis, **DON'T** immediately conclude that the variable has no association with the response.

-   There may be a linear association that is just not strong enough to detect given your data, or there may be a non-linear association.
:::
::::

## Recap

-   Reviewed model comparison
-   Introduced inference for multiple linear regression

## Next class

-   Exam 01 review
