---
title: "Logistic Regression: Prediction"
author: "Prof. Maria Tackett"
date: "2025-03-25"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 210 - Spring 2025](https://sta210-sp25.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 18-logistic-prediction-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE,
	echo = FALSE
)
```

## Announcements

-   Project presentations in March 31 lab

-   Statistics experience due April 15

<br>

::: question
Drop-in Peer Advising will be happening Wednesday, March 26, from 6:30-8:30 PM in Bostock Lounge for students to stop by with any questions they have around STA/CS/MATH classes, major pathways, or other general advice since shopping carts just opened.
:::

## Fall 2025 STA classes {.midi}

-   **STA 221:** Regression Analysis - Theory and Applications (Math 216/218/221)

-   **STA 240:** Probability for Statistical Inference, Modeling, and Data Analysis (or STA 230 or STA 231)

-   **STA 322:** Study Design: Design of Surveys and Causal Studies

-   **STA 323:** Statistical Computing

-   **STA 325:** Machine Learning and Data Mining (need STA 230/231/240)

-   **STA 332:** Statistical Inference (need STA 230/231/240 and Math 202/212/219)

-   **STA 344:** Introduction to Statistical Modeling of Spatial and Time Series data (need STA 230/231/240)

## Computational set up

```{r}
#| echo: true
#| warning: false
#| message: false


library(tidyverse)
library(tidymodels)
library(pROC)       # make ROC curves
library(knitr)
library(kableExtra)
library(patchwork)

# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

## Topics

-   Calculating predicted probabilities from the logistic regression model

-   Using predicted probabilities to classify observations

-   Make decisions and assess model performance using

    -   Confusion matrix
    -   ROC curve

## Data: Risk of coronary heart disease

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `high_risk`: 1 = High risk of having heart disease in next 10 years, 0 = Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `totChol`: Total cholesterol (in mg/dL)

-   `currentSmoker`: 0 = nonsmoker; 1 = smoker

## Univariate EDA

```{r}
#| echo: false
heart_disease <- read_csv("data/framingham.csv") |>
  select(age, totChol, TenYearCHD, currentSmoker) |>
  drop_na() |>
  mutate(high_risk = as_factor(TenYearCHD), 
         currentSmoker = as_factor(currentSmoker))
```

```{r}
#| echo: false

p1 <- ggplot(data = heart_disease, aes(x = high_risk)) + 
  geom_bar(fill = "darkcyan", color = 'black') + 
  labs(x = "High risk")

p2 <- ggplot(data = heart_disease, aes(x = age)) + 
  geom_histogram(fill = "darkcyan",color = 'black') + 
  labs(x = "Age")
p3 <- ggplot(data = heart_disease, aes(x = totChol)) + 
  geom_histogram(fill = "darkcyan", color = 'black') + 
  labs(x = "Total Cholesterol")

p4 <- ggplot(data = heart_disease, aes(x = currentSmoker)) + 
  geom_bar(fill = "darkcyan", color = 'black') + 
  labs(x = "Current smoker")

(p1 + p2) / (p3 + p4)
```

## Bivariate EDA

```{r}
#| echo: false

p5 <- ggplot(data = heart_disease, aes(x = high_risk, y = age)) + 
  geom_boxplot(fill = "steelblue") +
  labs(x = "High risk",
       y = "Age") + 
  coord_flip()

p6 <- ggplot(data = heart_disease, aes(x = currentSmoker, fill = high_risk)) + 
  geom_bar(position = "fill") +
  labs(x = "Current smoker", 
       fill = "High risk") + 
  theme(legend.position = "bottom")

p7 <- ggplot(data = heart_disease, aes(x = high_risk, y = totChol)) + 
  geom_boxplot(fill = "steelblue") +
  labs(x = "High risk",
       y = "Total cholesterol") + 
  coord_flip()

p6 + (p5 / p7) 

```

## Modeling risk of coronary heart disease

```{r}
heart_disease_fit <- glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")

tidy(heart_disease_fit) |> 
  kable(digits = 3)
```

::: question
Interpret the effect of `age` in terms of the odds of being high risk of heart disease.
:::

## Prediction and classification

::: incremental
-   We are often interested in using the model to classify observations, i.e., predict whether a given observation will have a 1 or 0 response

-   For each observation

    -   Use the logistic regression model to calculate the predicted log-odds the response for the $i^{th}$ observation is 1
    -   Use the log-odds to calculate the predicted probability the $i^{th}$ observation is 1
    -   Then, use the predicted probability to classify the observation as having a 1 or 0 response using some predefined threshold
:::

## Augmented data frame

```{r}
#| echo: true
augment(heart_disease_fit)
```

## Predicted log-odds

```{r}
#| echo: true
heart_disease_aug <- augment(heart_disease_fit)
```

```{r}
#| echo: false
heart_disease_aug |> 
  select(.fitted) |> 
  slice(1:5)
```

. . .

**Observation 1**

$$
\log(\hat{\text{odds}}) = \log\Big(\frac{\hat{p}}{1- \hat{p}}\Big) = -3.06
$$

## Predicted odds

```{r}
#| echo: false
heart_disease_aug |> 
  select(.fitted) |> 
  slice(1:5)
```

. . .

**Observation 1**

$$
\hat{\text{odds}} = \frac{\hat{p}}{1- \hat{p}} = \exp\{-3.06\} = 0.0469
$$

## Predicted probability

```{r}
#| echo: false
heart_disease_aug |> 
  select(.fitted) |> 
  slice(1:5)
```

. . .

**Observation 1**

$$
\text{predicted prob.} = \hat{p} = \frac{\hat{\text{odds}}}{1+\hat{\text{odds}}} = \frac{\exp\{-3.06\}}{1 + \exp\{-3.06\}}= 0.045
$$

. . .

::: question
Would you classify this individual as high risk $(\hat{y} = 1)$ or not high risk $(\hat{y} = 0)$?
:::

## Another individual

```{r}
#| echo: false
heart_disease_aug |> 
  select(.fitted) |> 
  slice(1:5)
```

. . .

**Observation 4**

$$
\text{predicted prob.} = \hat{p} = \frac{\hat{\text{odds}}}{1+\hat{\text{odds}}} = \frac{\exp\{-0.751\}}{1 + \exp\{-0.751\}}= 0.321
$$

. . .

::: question
Would you classify this individual as high risk $(\hat{y} = 1)$ or not high risk $(\hat{y} = 0)$?
:::

## Predicted probabilities in R

We can calculate predicted probabilities using the `predict.glm()` function. Use `type = "response"` to get probabilities.[^1]

[^1]: The default is `type = "link"`, which produces the predicted log-odds.

<br>

. . .

```{r}
#| eval: false
#| echo: true
predict.glm(heart_disease_fit, type = "response")
```

. . .

**Predicted probabilities for Observations 1 -5**

```{r}
#| echo: false
predict.glm(heart_disease_fit, type = "response")[1:5]
```

## Predictions in R

```{r}
#| echo: true

pred_prob <- predict.glm(heart_disease_fit, type = "response")

heart_disease_aug <- heart_disease_aug |> 
  bind_cols(pred_prob = pred_prob)
```

. . .

```{r}
#| echo: false

heart_disease_aug  |> 
  select(high_risk, .fitted, pred_prob) |>
  slice(1:5)
```

## Classifying observations

::: question
You would like to determine a threshold for classifying individuals as high risk or not high risk.

What considerations would you make in determining the threshold?
:::

## Classify using 0.5 as threshold

We can use a threshold of 0.5 to classify observations.

-   If $\hat{p} > 0.5$, classify as 1

-   If $\hat{p} \leq 0.5$, classify as 0

. . .

```{r}
heart_disease_aug <- heart_disease_aug |>
  mutate(pred_class = factor(if_else(pred_prob > 0.5, 1, 0)))
```

. . .

```{r}
#| echo: false
heart_disease_aug |>
  select(high_risk, .fitted, pred_prob, pred_class) |>
  slice(1:5)
```

## Confusion matrix

A **confusion matrix** is a $2 \times 2$ table that compares the predicted and actual classes. We can produce this matrix using the `conf_mat()` function in the **yardstick** package (part of tidymodels).

<br>

. . .

```{r}
#| echo: true
heart_disease_aug |>
  conf_mat(high_risk, pred_class) 
```

## Visualize confusion matrix

```{r}
#| echo: true
heart_conf_mat <- heart_disease_aug |>
  conf_mat(high_risk, pred_class)

autoplot(heart_conf_mat, type = "heatmap")
```

## Using the confusion matrix

::: center
```{r}
#| echo: false

heart_disease_aug |>
  conf_mat(high_risk, pred_class) 
```
:::

<br>

. . .

The **accuracy** of this model with a classification threshold of 0.5 is

$$
\text{accuracy} = \frac{3553 + 0}{3553 + 635 + 2 + 0} = 0.848
$$

## Using the confusion matrix

::: center
```{r}
#| echo: false

heart_disease_aug |>
  conf_mat(high_risk, pred_class) 
```
:::

<br>

. . .

The **misclassification rate** of this model with a threshold of 0.5 is

$$
\text{misclassification} = \frac{635 + 2}{3553 + 635 + 2 + 0} = 0.152
$$

## Using the confusion matrix

::: center
```{r}
#| echo: false

heart_disease_aug |>
  conf_mat(high_risk, pred_class) 
```
:::

<br>

Accuracy is 0.848 and the misclassification rate is 0.152.

. . .

::: question
-   What is the limitation of solely relying on accuracy and misclassification to assess the model performance?

-   What is the limitation of using a single confusion matrix to assess the model performance?
:::

# Sensitivity and specificity

## True/false positive/negative {.midi}

|   | <u>Not</u> high risk $(y_i = 0)$ | High risk $(y_i = 1)$ |
|----|----|----|
| **Classified** <u>**not**</u> **high risk** $(\hat{p}_i \leq \text{threshold})$ | True negative (TN) | False negative (FN) |
| **Classified high risk** $(\hat{p}_i > \text{threshold})$ | False positive (FP) | True positive (TP) |

<br>

-   $\text{accuracy} = \frac{TN + TP}{TN + TP + FN + FP}$

-   $\text{misclassification} = \frac{FN + FP}{TN+ TP + FN + FP}$

## False negative rate {.midi}

|   | Not high risk $(y_i = 0)$ | High risk $(y_i = 1)$ |
|----|----|----|
| **Classified not high risk** $(\hat{p}_i \leq \text{threshold})$ | True negative (TN) | False negative (FN) |
| **Classified high risk** $(\hat{p}_i > \text{threshold})$ | False positive (FP) | True positive (TP) |

<br>

. . .

**False negative rate**: Proportion of actual positives that were classified as negatives

-   P(classified not high risk \| high risk) = $\frac{FN}{TP + FN}$

## False positive rate {.midi}

|   | Not high risk $(y_i = 0)$ | High risk $(y_i = 1)$ |
|----|----|----|
| **Classified not high risk** $(\hat{p}_i \leq \text{threshold})$ | True negative (TN) | False negative (FN) |
| **Classified high risk** $(\hat{p}_i > \text{threshold})$ | False positive (FP) | True positive (TP) |

<br>

. . .

**False positive rate**: Proportion of actual negatives that were classified as positives

-   P(classified high risk \| not high risk) = $\frac{FP}{TN + FP}$

## Sensitivity {.midi}

|   | Not high risk $(y_i = 0)$ | High risk $(y_i = 1)$ |
|----|----|----|
| **Classified not high risk** $(\hat{p}_i \leq \text{threshold})$ | True negative (TN) | False negative (FN) |
| **Classified high risk** $(\hat{p}_i > \text{threshold})$ | False positive (FP) | True positive (TP) |

<br>

. . .

**Sensitivity**: Proportion of actual positives that were correctly classified as positive

-   Also known as *true positive rate* (TPR) and *recall*

-   **P(classified high risk \| high risk) = 1 − False negative rate**

## Specificity {.midi}

|   | Not high risk $(y_i = 0)$ | High risk $(y_i = 1)$ |
|----|----|----|
| **Classified not high risk** $(\hat{p}_i \leq \text{threshold})$ | True negative (TN) | False negative (FN) |
| **Classified high risk** $(\hat{p}_i > \text{threshold})$ | False positive (FP) | True positive (TP) |

<br>

. . .

**Specificity**: Proportion of actual negatives that were correctly classified as negative

-   **P(classified not high risk \| not high risk) = 1 − False positive rate**

## Practice

::: center
```{r}
#| echo: false

heart_disease_aug |>
  conf_mat(high_risk, pred_class) 
```
:::

::: question
Calculate the

-   False negative rate
-   False positive rate
-   Sensitivity
-   Specificity
:::

## Using metrics to select model and threshold {.midi}

+----------------------------------+---------------------------------------------------------------------+
| Metric                           | Guidance for use                                                    |
+==================================+=====================================================================+
| Accuracy                         | For balanced data, use only in combination with other metrics.      |
|                                  |                                                                     |
|                                  | Avoid using for imbalanced data.                                    |
+----------------------------------+---------------------------------------------------------------------+
| Sensitivity (true positive rate) | Use when false negatives are more "expensive" than false positives. |
+----------------------------------+---------------------------------------------------------------------+
| False positive rate              | Use when false positives are more "expensive" than false negatives. |
+----------------------------------+---------------------------------------------------------------------+
| Precision = $\frac{TP}{TP + FP}$ | Use when it's important for positive predictions to be accurate.    |
+----------------------------------+---------------------------------------------------------------------+

::: small
This table is a modification of work created and shared by Google in the [Google Machine Learning Crash Course](#0).
:::

## Choosing a classification threshold

::: question
A doctor plans to use your model to determine which patients are high risk for heart disease. The doctor will recommend a treatment plan for high risk patients.

-   Would you want sensitivity to be high or low? What about specificity?

-   What are the trade-offs associated with each decision?
:::

## ROC curve

So far the model assessment has depended on the model and selected threshold. The **receiver operating characteristic (ROC) curve** allows us to assess the model performance across a range of thresholds.

. . .

::::: columns
::: {.column width="65%"}
```{r}
#| echo: false

# calculate sensitivity and specificity at each threshold
roc_curve_data <- heart_disease_aug |>
  roc_curve(high_risk, .fitted, 
            event_level = "second") 

# plot roc curve
autoplot(roc_curve_data)
```
:::

::: {.column width="35%"}
-   x-axis: 1 - Specificity (False positive rate)

-   y-axis: Sensitivity (True positive rate)

Which corner of the plot indicates the best model performance?
:::
:::::

## ROC curve

```{r}
#| echo: false
#| out-width: 100%

heart_disease_aug |>
  roc_curve(high_risk, .fitted, 
            event_level = "second")  |>
  autoplot() +
  annotate("point", x = 0, y = 1, color = "#5B888C") +
  annotate("point", x = 0, y = 1, color = "#5B888C", size = 3, shape = "circle open") +
  annotate(
    "label", x = 0.02, y = 0.99, label = "All high risk classified as high risk,\nall not high risk classified as not high risk", hjust = 0, color = "#5B888C", fontface = "bold", vjust = 1, fill = "white"
  ) +
  annotate("point", x = 1, y = 0, color = "#8F2D56") +
  annotate("point", x = 1, y = 0, color = "#8F2D56", size = 3, shape = "circle open") +
  annotate(
    "label", x = 0.98, y = 0.01, label = "All high risk classified as not high risk,\nall not high risk classified as high risk", hjust = 1, color = "#8F2D56", fontface = "bold", vjust = 0, fill = "white"
  ) +
  annotate(
    "segment", color = "#5b708c", x = 0, y = 0, xend = 1, yend = 1, size = 2, alpha = 0.6
  ) +
  annotate(
    "label", x = 0.58, y = 0.5, label = "True positive rate\n= false positive rate", hjust = 0, color = "#5b708c", fontface = "bold", fill = "white"
  )
```

## ROC curve in R {.midi}

```{r}
#| echo: true
#| eval: false


# calculate sensitivity and specificity at each threshold
roc_curve_data <- heart_disease_aug |>
  roc_curve(high_risk, .fitted, 
            event_level = "second") 

# plot roc curve
autoplot(roc_curve_data)
```

## ROC curve in R

::::: columns
::: {.column width="40%"}
```{r}
#| echo: false

roc_curve_data |>
  slice(1:20)

```
:::

::: {.column width="60%"}
```{r}
#| echo: false

autoplot(roc_curve_data)
```
:::
:::::

## Area under the curve

The **area under the curve (AUC)** can be used to assess how well the logistic model fits the data

-   AUC=0.5: model is a very bad fit (no better than a coin flip)

-   AUC close to 1: model is a good fit

. . .

```{r}
#| echo: true
heart_disease_aug |>
  roc_auc(high_risk, .fitted,
    event_level = "second"
  )
```

## Recap

-   Calculated predicted probabilities from the logistic regression model

-   Used predicted probabilities to classify observations

-   Made decisions and assessed model performance using

    -   Confusion matrix
    -   ROC curve

## Further reading

[Classification](https://developers.google.com/machine-learning/crash-course/classification) module in Google Machine Learning Crash Course
