---
title: "Logistic Regression: Model comparison"
author: "Prof. Maria Tackett"
date: "2025-03-27"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 221 - Spring 2025](https://sta221-sp25.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 19-logistic-compare-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Announcements

-   Project presentations in March 31 lab

-   Statistics experience due April 15

## Topics

Comparing logistic regression models using

-   Drop-in-deviance test

-   AIC

-   BIC

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(knitr)
library(kableExtra)  # for table embellishments
library(Stat2Data)   # for empirical logit

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Data

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `high_risk`:

    -   1: High risk of having heart disease in next 10 years
    -   0: Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `education`: 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

-   `currentSmoker`: 0 = nonsmoker, 1 = smoker

-   `totChol`: Total cholesterol (in mg/dL)

## Data

```{r}
#| echo: false
heart_disease <- read_csv(here::here("slides", "data/framingham.csv")) |>
  select(age, education, TenYearCHD, totChol, currentSmoker) |>
  drop_na() |> #consider the limitations of doing this
  mutate(
    high_risk = as.factor(TenYearCHD),
    education = as.factor(education),
    currentSmoker = as.factor(currentSmoker)
  )

heart_disease
```

## Modeling risk of coronary heart disease

Using `age`, `totChol`, and `currentSmoker`

```{r}
#| echo: false
high_risk_fit <- glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")

tidy(high_risk_fit, conf.int = TRUE) |> 
  kable(format = "markdown", digits = 3)
```

## Review: ROC Curve + Model fit

```{r}
high_risk_aug <- augment(high_risk_fit)

roc_curve_data <- high_risk_aug |>
  roc_curve(high_risk, .fitted, event_level = "second")

# plot roc curve
autoplot(roc_curve_data)

#calculate AUC
high_risk_aug |>
  roc_auc(high_risk, .fitted, event_level = "second")
```

## Review: Classification

We will use a threshold of 0.2 to classify observations

```{r}
#compute predicted probabilities
pred_prob <- predict.glm(high_risk_fit, type = "response")
high_risk_aug <- high_risk_aug |> 
  bind_cols(pred_prob = pred_prob) |>
  mutate(pred_class = factor(if_else(pred_prob > 0.2, 1, 0)))

# compute confusion matrix
high_risk_aug |>
  conf_mat(high_risk, pred_class) |>
  autoplot(type = "heatmap")
```

## Review: Classification  {.midi}

```{r}
#| echo: false

high_risk_aug |>
  conf_mat(high_risk, pred_class) |>
  autoplot(type = "heatmap")
```

::: question
1.  Compute the misclassification rate.

2.  Compute sensitivity and explain what it means in the context of this data.

3.  Compute specificity and explain what it means in the context of this data.
:::

# Drop-in-deviance test

## Which model do we choose? 

```{r}
#| echo: false

high_risk_fit2 <- glm(high_risk ~ age + totChol + currentSmoker + education, 
              data = heart_disease, family = "binomial")
```

::::: columns
::: {.column width="50%"}
<center>**Model 1**</center>

```{r}
#| echo: false

tidy(high_risk_fit) |> 
  select(term, estimate) |>
  kable(digits = 3)

```
:::

::: {.column width="50%"}
<center>**Model 2** </center>

```{r}
#| echo: false

tidy(high_risk_fit2) |> 
  select(term, estimate) |>
  kable(digits = 3)
```
:::
:::::

## Log likelihood

$$
\log L = \sum\limits_{i=1}^n[y_i \log(\hat{p}_i) + (1 - y_i)\log(1 - \hat{p}_i)]
$$

-   Measure of how well the model fits the data

-   Higher values of $\log L$ are better

-   **Deviance** = $-2 \log L$

    -   $-2 \log L$ follows a $\chi^2$ distribution with $n - p - 1$ degrees of freedom

## Comparing nested models

-   Suppose there are two nested models:

    -   Reduced Model includes predictors $x_1, \ldots, x_q$
    -   Full Model includes predictors $x_1, \ldots, x_q, x_{q+1}, \ldots, x_p$

-   We want to test the hypotheses

    $$
    \begin{aligned}
    H_0&: \beta_{q+1} = \dots = \beta_p = 0 \\
    H_a&: \text{ at least one }\beta_j \text{ is not } 0
    \end{aligned}
    $$

-   To do so, we will use the **Drop-in-deviance test**, also known as the Nested Likelihood Ratio test

## Drop-in-deviance test

**Hypotheses:**

$$
\begin{aligned}
H_0&: \beta_{q+1} = \dots = \beta_p = 0 \\
H_a&: \text{ at least one }\beta_j \text{ is not } 0
\end{aligned}
$$

. . .

**Test Statistic:** $$\begin{aligned}G &= \text{Deviance}_{reduced} - \text{Deviance}_{full} \\
&= (-2 \log L_{reduced}) - (-2 \log L_{full})\end{aligned}$$

. . .

**P-value:** $P(\chi^2 > G)$, calculated using a $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters in the full and reduced models

## $\chi^2$ distribution

```{r}
#| echo: false
#| fig-height: 6

x <- seq(from =0, to = 10, length = 100)

# Evaluate the densities
y_1 <- dchisq(x, 1)
y_2 <- dchisq(x,2)
y_3 <- dchisq(x,3)
y_4 <- dchisq(x,5)

# Plot the densities
plot(x, y_1, col = 1, type = "l", ylab="",lwd=3, ylim = c(0, 0.5), 
     main  = "Chi-square Distribution")
lines(x,y_2, col = 2,lwd=3)
lines(x, y_3, col = 3,lwd=3)
lines(x, y_4, col = 4,lwd=3)

# Add the legend
legend("topright",
       c("df = 1", "df = 2 ", "df = 3", "df = 5"), 
       col = c(1, 2, 3, 4), lty = 1)
```

## Should we add `education` to the model?

First model, reduced:

```{r}

high_risk_fit_reduced <- glm(high_risk ~ age + totChol + currentSmoker,
                             data = heart_disease, family = "binomial")
```

<br>

. . .

Second model, full:

```{r}
#| code-line-numbers: "3"

high_risk_fit_full <- glm(high_risk ~ age + totChol +
                            currentSmoker + education, 
              data = heart_disease, family = "binomial")
```

::: question
Write the null and alternative hypotheses in words and mathematical notation.
:::

## Should we add `education` to the model?

Calculate deviance for each model:

```{r}
(dev_reduced <- glance(high_risk_fit_reduced)$deviance)

(dev_full <- glance(high_risk_fit_full)$deviance)
```

. . .

Drop-in-deviance test statistic:

```{r}
(test_stat <- dev_reduced - dev_full)
```

## Should we add `education` to the model?

Calculate the p-value using a `pchisq()`, with degrees of freedom equal to the number of new model terms in the second model:

```{r}
pchisq(test_stat, 3, lower.tail = FALSE) 
```

<br>

. . .

::: question
What is your conclusion?
:::

## Drop-in-Deviance test in R {.midi}

-   We can use the **`anova`** function to conduct this test

-   Add **`test = "Chisq"`** to conduct the drop-in-deviance test

. . .

```{r}
anova(high_risk_fit_reduced, high_risk_fit_full, test = "Chisq") |>
  tidy() |> kable(digits = 3)
```

# Model selection using AIC and BIC

## AIC & BIC

Estimators of prediction error and *relative* quality of models:

. . .

**Akaike's Information Criterion (AIC)**[^1]: $$AIC = -2\log L + 2(p+1)$$

[^1]: Akaike, Hirotugu. ["A new look at the statistical model identification."](https://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=1100705) *IEEE transactions on automatic control* 19.6 (1974): 716-723.

. . .

**Schwarz's Bayesian Information Criterion (BIC)**[^2]: $$ BIC = -2\log L + \log(n)\times(p+1)$$

[^2]: Schwarz, Gideon. ["Estimating the dimension of a model."](https://projecteuclid.org/journalArticle/Download?urlId=10.1214%2Faos%2F1176344136) *The annals of statistics* (1978): 461-464.

## AIC & BIC

$$
\begin{aligned} 
& AIC = \color{blue}{-2\log L}  \color{black}{+ 2(p+1)} \\
& BIC = \color{blue}{-2\log L}  + \color{black}{\log(n)\times(p+1)}
\end{aligned}
$$

. . .

<br>

First Term: Decreases as *p* increases

## AIC & BIC

$$
\begin{aligned} & AIC = -2\log L  + \color{blue}{2(p+1)} \\
& BIC = -2\log L + \color{blue}{\log(n)\times(p+1)} 
\end{aligned}
$$

<br>

Second term: Increases as *p* increases

## Using AIC & BIC

$$
\begin{aligned} & AIC = -2\log L  + \color{red}{2(p+1)} \\
& BIC = -2 \log L  + \color{red}{\log(n)\times(p+1)} 
\end{aligned}
$$

-   Choose model with the smaller value of AIC or BIC

-   If $n \geq 8$, the **penalty** for BIC is larger than that of AIC, so BIC tends to favor *more parsimonious* models (i.e. models with fewer terms)

## AIC from the `glance()` function

Let's look at the AIC for the model that includes `age`, `totChol`, and `currentSmoker`

```{r}
glance(high_risk_fit)$AIC
```

<br>

. . .

**Calculating AIC**

```{r}
- 2 * glance(high_risk_fit)$logLik + 2 * (3 + 1)
```

## Comparing the models using AIC

Let's compare the full and reduced models using AIC.

```{r}
glance(high_risk_fit_reduced)$AIC
glance(high_risk_fit_full)$AIC
```

<br>

::: question
Based on AIC, which model would you choose?
:::

## Comparing the models using BIC

Let's compare the full and reduced models using BIC

```{r echo = T}
glance(high_risk_fit_reduced)$BIC
glance(high_risk_fit_full)$BIC
```

<br>

::: question
Based on BIC, which model would you choose?
:::

# Application exercise

::: appex
📋 [sta210-sp25.netlify.app/ae/ae-10-logistic-compare.html](../ae/ae-10-logistic-compare.html)
:::
