---
title: "Cross validation"
author: "Prof. Maria Tackett"
date: "2025-04-10"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 210 - Spring 2025](https://sta210-sp25.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 23-cross-validation-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

## Announcements

-   Exam 02 - April 17 (same format as Exam 01)

    -   Lecture videos available

-   Next project milestone: Draft and peer review in April 21 lab

-   Statistics experience due April 15

## Topics

::: nonincremental
-   Cross validation
:::

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(patchwork)
library(knitr)
library(kableExtra)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

## Data: Sesame Street

Today's data comes from [an experiment](https://files.eric.ed.gov/fulltext/ED122799.pdf) by the Educational Testing Service to test the effectiveness of the children's program *Sesame Street*. *Sesame Street* is an educational program designed to teach young children basic educational skills such as counting and the alphabet

As part of the experiment, children were assigned to one of two groups: those who were encouraged to watch the program and those who were not.

The show is only effective if children watch it, so we want to understand what effect the encouragement had on the frequency children watched the program.

## Variables

**Response:**

-   `viewcat`
    -   1: rarely watched show
    -   2: watched once or twice a week
    -   3: watched three to five times a week
    -   4: watched show on average more than five times a week

**Predictors:**

-   `age`: child's age in months
-   `viewenc`: 1: encouraged to watch, 0: not encouraged

# Application exercise

::: appex
📋 <https://sta210-sp25.netlify.app/ae/ae-11-multinomial.html>
:::

# Cross validation

## Spending our data

-   We have already established that the idea of data spending where the test set was recommended for obtaining an unbiased estimate of performance.
-   However, we usually need to understand the effectiveness of the model [*before*]{.underline} *using the test set*.
-   Typically we can't decide on *which* final model to take to the test set without making model assessments.
-   **Remedy:** Resampling to make model assessments on training data in a way that can generalize to new data.

## Resampling for model assessment

**Resampling is only conducted on the** <u>**training**</u> **set**. The test set is not involved. For each iteration of resampling, the data are partitioned into two subsamples:

-   The model is fit with the **analysis set**. Model fit statistics such as $R^2_{Adj}$, AIC, BIC, and AUC are calculated based on this fit.
-   The model is evaluated with the **assessment set**.

## Resampling for model assessment

![](images/23/resampling.svg){fig-align="center"}

<br>

Image source: Kuhn and Silge. [Tidy modeling with R](https://www.tmwr.org/).

## Analysis and assessment sets

-   Analysis set is analogous to training set.
-   Assessment set is analogous to test set.
-   The terms *analysis* and *assessment* avoids confusion with initial split of the data.
-   These data sets are mutually exclusive.

## Cross validation

More specifically, **v-fold cross validation** -- commonly used resampling technique:

-   Randomly split your **training** **data** into ***v*** partitions
-   Use ***v-1*** partitions for analysis, and the remaining 1 partition for analysis (model fit + model fit statistics)
-   Repeat ***v*** times, updating which partition is used for assessment each time

. . .

Let's give an example where `v = 3`...

```{r}
#| echo: false

sesame <- read_csv("https://bit.ly/sesame-street-data")

# mean-center relevant continuous variables, make categorical variables factors
sesame <- sesame |> 
  mutate(viewcat = factor(viewcat), 
         site = factor(site), 
         prenumbCent = prenumb - mean(prenumb), 
         preletCent = prelet - mean(prelet), 
         ageCent = age - mean(age),
         viewenc = factor(if_else(viewenc == 1, "1", "0"))
  )

```

## To get started...

**Split data into training and test sets**

```{r}
#| echo: true
set.seed(345)

sesame_split <- initial_split(sesame)
sesame_train <- training(sesame_split)
sesame_test <- testing(sesame_split)
```

## To get started... {.midi}

**Specify model**

```{r}
#| echo: true
sesame_spec <- multinom_reg()
```

<br>

. . .

```{r}
sesame_spec
```

<br>

. . .

**Note:** Use `linear_reg()` or `logistic_reg()` for linear or logistic models, respectively.

## To get started...

**Create workflow**

```{r}
#| echo: true
sesame_wflow1 <- workflow() |>
  add_model(sesame_spec) |>
  add_formula(viewcat ~ ageCent + viewenc + site)
```

. . .

```{r}
sesame_wflow1
```

## Cross validation, step 1

Randomly split your **training** **data** into 3 partitions:

<br>

![](images/23/three-CV.svg){fig-align="center"}

## Tips: Split training data

```{r}
#| echo: true
folds <- vfold_cv(sesame_train, v = 3)
folds
```

## Cross validation, steps 2 and 3

::: nonincremental
-   Use *v-1* partitions for analysis, and the remaining 1 partition for assessment
-   Repeat *v* times, updating which partition is used for assessment each time
:::

![](images/23/three-CV-iter.svg){fig-align="center"}

## Sesame Street: Fit resamples {.midi}

```{r}
sesame_fit_rs1 <- sesame_wflow1 |>
  fit_resamples(resamples = folds,
               metrics = metric_set(accuracy, roc_auc))

sesame_fit_rs1
```

## Cross validation, now what?

-   We've fit a bunch of models
-   Now it's time to use them to collect metrics (e.g., AUC, ) on each model and use them to evaluate model fit and how it varies across folds

## Collect metrics from CV

```{r}
# Produces summary across all CV
collect_metrics(sesame_fit_rs1)
```

<br>

Note: These are calculated using the *assessment* data

## Deeper look into results

```{r}
cv_metrics1 <- collect_metrics(sesame_fit_rs1, summarize = FALSE) 

cv_metrics1
```

## Better presentation of results

```{r}
cv_metrics1 |>
  mutate(.estimate = round(.estimate, 3)) |>
  pivot_wider(id_cols = id, names_from = .metric, values_from = .estimate) |>
  kable(col.names = c("Fold", "Accuracy", "AUC"))
```

## Cross validation in practice

::: incremental
-   To illustrate how CV works, we used `v = 3`:

    ::: nonincremental
    -   Analysis sets are 2/3 of the training set
    -   Each assessment set is a distinct 1/3
    -   The final resampling estimate of performance averages each of the 3 replicates
    :::

-   This was useful for illustrative purposes, but `v` is often 5 or 10; we generally prefer 10-fold cross-validation as a default
:::

## Example modeling workflow 

-   Exploratory data analysis

-   Using training data...

    -   Fit and evaluate candidate model using cross validation

    -   Select the best fit model

    -   Check model conditions and diagnostics

    -   Repeat as needed until you've landed on final model

-   Evaluate the final model performance using the test set

## Data analysis workflow

![Source: [*R for Data Science*](https://r4ds.hadley.nz/) with additions from *The Art of Statistics: How to Learn from Data*.](images/02/data-analysis-life-cycle.png){alt="Source: R for Data Science with additions from The Art of Statistics: How to Learn from Data."}
