---
title: "Lab 01: Ice duration and air temperature in Madison, WI"
subtitle: "Simulation-based inference for the slope"
execute:
  freeze: auto
  echo: true
  eval: false
  warning: false
  message: false
---

::: callout-important
This lab is due on **Thursday, January 30 at 11:59pm.**
:::

# Introduction

In today's lab, you'll use linear regression to analyze the relationship between air temperature and ice duration for two lakes in Wisconsin.

## Learning goals

By the end of the lab you will...

-   Fit and interpret a simple linear regression model in R.
-   Be able to use simulation-based inference to draw conclusions about the slope.
-   Continue developing a workflow for reproducible data analysis.

# Getting started

-   Go to the [sta210-sp25](https://www.github.com/sta210-sp25) organization on GitHub. Click on the repo with the prefix **lab-01-**. It contains the starter documents you need to complete the lab.

-   Clone the repo and start a new project in RStudio. See the [Lab 00 instructions](lab-00.qmd) for details on cloning a repo, starting a new project in R, and configuring git.

# Packages

The following packages are used in the lab:

```{r load-packages}
#| eval: true
library(tidyverse)
library(tidymodels)
library(knitr)
library(lterdatasampler) 

# add other packages as needed
```

# Data: Ice cover and air temperature

The datasets `wi-icecover.csv` and `wi-air-temperature.csv` contain information about ice cover and air temperature, respectively, at Lake Monona and Lake Mendota (both in Madison, Wisconsin) for days in 1886 through 2019. The data were obtained from the [`ntl_icecover`](https://lter.github.io/lterdatasampler/reference/ntl_icecover.html) and [`ntl_airtemp`](https://lter.github.io/lterdatasampler/reference/ntl_airtemp.html) data frames in the [lterdatasampler](https://lter.github.io/lterdatasampler/) R package. They were originally collected by the [US Long Term Ecological Research program (LTER) Network](https://lternet.edu/).

The primary variables for this analysis are

-   `year`: year of observation

-   `lakeid`: lake name

-   `ice_duration`: number of days between the freeze and ice breakup dates of each lake

-   `air_temp_avg`: yearly average air temperature in Madison, WI (degrees Celsius)

```{r}
#| message: false
icecover <- read_csv("data/wi-icecover.csv")
airtemp <- read_csv("data/wi-air-temperature.csv")
```

# Exercises

**Goal**: The goal of this analysis is to use linear regression understand the association between average air temperature and ice duration for two lakes in Madison, Wisconsin that freeze for a portion of the year. Because ice cover is impacted by various environmental factors, researchers are interested in examining the association between these two factors to better understand how changing air temperature is impacting the ice duration.

------------------------------------------------------------------------

Write all code and narrative in your Quarto file. Write all narrative in complete sentences. Throughout the assignment, you should periodically **render** your Quarto document to produce the updated PDF, **commit** the changes in the Git pane, and **push** the updated files to GitHub.

::: callout-tip
Make sure we can read all of your code in your PDF document. This means you will need to break up long lines of code. One way to help avoid long lines of code is is start a new line after every pipe (`|>`) and plus sign (`+`).
:::

### Exercise 1

Let's start by combining the data into a single data frame.

a.  Join `icecover` and `airtemp` to create a new data frame that contains both the ice temperature and average air temperature. Your new data frame will have one row per lake per year.

b.  How many observations (rows) are in the new data frame? How many variables (columns)?

c.  We will also consider the change of ice duration over time. Add a new variable to the combined data frame that is the number of years since 1886 (the first year in the data set).

::: callout-important
You will use the combined data for the remainder of the lab.
:::

### Exercise 2

a.  Make a histogram to visualize the distribution of the response variable `ice_duration` and calculate summary statistics for the center and spread of the distribution. Include informative axis labels and an informative title on the visualization.

b.  Use the visualization and summary statistics to describe the distribution of `ice_duration`. Include the shape, center, spread, and potential presence of outliers in the description.

### Exercise 3

a.  Now visualize the distribution of `ice_duration` using a different type of plot.

::: callout-tip
See the [Data visualization with ggplot2 cheatsheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf) for inspiration.
:::

b.  What type of plot did you make? What is one feature of the distribution of ice duration that is more apparent in the plot from this exercise than the histogram? What is one feature of the distribution that is more apparent in the histogram than the plot in this exercise?

::: render-commit-push
This is a good place to render, commit, and push changes to your remote lab repo on GitHub. Click the checkbox next to each file in the Git pane to stage the updates you've made, write an informative commit message (e.g., "Completed Exercises 1 - 3"), and push. After you push the changes, the Git pane in RStudio should be empty.
:::

### Exercise 4

Let's look at the change in ice duration over time.

a.  Visualize ice duration over time for each lake.

b.  Use the visualization to describe how ice duration has generally changed over time.

c.  Based on the visualization, do you think `lakeid` is a useful predictor of ice duration? Briefly explain why or why not.

### Exercise 5

a.  Fit the linear regression model using years since 1886 to understand variability in ice duration. Neatly display the results using 3 digits.

<!-- -->

c.  Interpret the slope in the context of the data.

d.  Does the intercept have a meaningful interpretation in this context? If so, interpret the intercept in the context of the data. Otherwise, explain why not.

### Exercise 6

Conduct a permutation test for the slope to assess whether there is sufficient evidence of a change in ice duration over time. In your response:

-   State the null and alternative hypotheses in words and mathematical notation
-   Show all relevant code and output used to conduct the test. Use `set.seed(2025)` and 1000 iterations to construct the appropriate distribution.
-   Visualize the simulated null distribution.
-   State the conclusion in the context of the data.

::: render-commit-push
This is a good place to render, commit, and push changes to your remote lab repo on GitHub. Click the checkbox next to each file in the Git pane to stage the updates you've made, write an informative commit message (e.g., "Completed Exercises 4 - 6"), and push. After you push the changes, the Git pane in RStudio should be empty.
:::

### Exercise 7

Researchers are interested in the effect of air temperature on ice duration, so let's explore that relationship.

a.  Fit the linear regression model average air temperature to understand variability in ice duration. Neatly display the results using 3 digits.

b.  Use bootstrapping to construct a 93% confidence interval for the slope using the same seed as before `2025`. In your response:

    -   First, set a seed for simulating reproducibly. Use the seed `210`.
    -   Show all relevant code and output used construct the bootstrap confidence intervals. Use `set.seed(2025)` and 1000 iterations to construct the appropriate distribution.
    -   Visualize the bootstrap distribution.

c.  Interpret the confidence interval for the slope in the context of the data.

### Exercise 8

We have seen evidence of changing ice duration over time and the effect of air temperature on ice duration. Now let's consider both of these effects in a single model.

a.  Fit a linear regression model using both years since 1886 and average air temperature to understand variability in ice duration. Neatly display the results using 3 digits.

b.  How does the slope of `air_temp_avg` in this model compare to the slope from the model in the previous exercise? Does the slope indicate a stronger, weaker, or same effect of air temperature on ice duration?

c.  Briefly explain why the slopes of air temperature differ between the two models.

::: render-commit-push
Render, commit, and push your final changes to your remote lab repo on GitHub. Click the checkbox next to each file in the Git pane to stage the updates you've made, write an informative commit message, and push. After you push the changes, the Git pane in RStudio should be empty.
:::

# Submission

In this class, we'll be submitting PDF documents to Gradescope.

::: callout-warning
Before you wrap up the assignment, make sure all documents are updated on your GitHub repo. We will be checking these to make sure you have been practicing how to commit and push changes.

Remember -- you must turn in a PDF file to the Gradescope page before the submission deadline for full credit.
:::

To submit your assignment:

-   Access Gradescope through the [STA 210 Canvas site](https://canvas.duke.edu/courses/47067).

-   Click on the assignment, and you’ll be prompted to submit it.

-   Mark the pages associated with each exercise. All of the pages of your lab should be associated with at least one question (i.e., should be “checked”).

-   Select the first page of your .PDF submission to be associated with the *“Workflow & formatting”* section.

# Grading (50 points)

<br>

| Component             | Points |
|:----------------------|:-------|
| Ex 1                  | 4      |
| Ex 2                  | 5      |
| Ex 3                  | 4      |
| Ex 4                  | 6      |
| Ex 5                  | 6      |
| Ex 6                  | 8      |
| Ex 7                  | 9      |
| Ex 8                  | 5      |
| Workflow & formatting | 3      |
