---
title: "Lab 02: Inference for linear regression using mathematical models"
subtitle: "Coffee ratings"
execute:
  freeze: auto
  echo: true
  eval: false
  warning: false
  message: false
---

::: callout-important
This lab is due on **Thursday, February 6 at 11:59pm.** To be considered on time, the following must be done by the due date:

-   Final `.qmd` and `.pdf` files pushed to your GitHub repo

-   Final `.pdf` file submitted on Gradescope
:::

# Introduction

In today's lab you will analyze data from over 1,000 different coffees to explore the relationship between a coffee's after taste and its overall quality. You will also meet your lab teams and start working collaboratively in GitHub.

## Learning goals

By the end of the lab you will...

-   use mathematical models to conduct inference for the slope.
-   assess conditions for linear regression.
-   develop a collaborative workflow in GitHub.

# Getting started

\[UPDATE with team instructions\]

# Packages

The following packages are used in the lab.

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
```

# Data: Coffee grades

The data set for this lab comes from the [Coffee Quality Database](https://github.com/jldbc/coffee-quality-database) and was obtained from the [#TidyTuesday GitHub repo](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-07-07/readme.md). It includes information about the origin, producer, measures of various characteristics, and the quality measure for over 1,000 coffees. The coffees can be reasonably be treated as a random sample.

This lab will focus on the following variables:

-   `aftertaste`: Aftertaste grade, 0 (worst aftertaste) - 10 (best aftertaste)
-   `total_cup_points`: Rating of overall quality, 0 (worst quality) - 10 (best quality)

[Click here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-07-07/readme.md) for the definitions of all variables in the data set. [Click here](http://www.scaa.org/?page=resources&d=cupping-protocols&mod=article_inline) for more details about how these measures are obtained.

```{r}
coffee <- read_csv("data/coffee-ratings.csv")
```

# Exercises

------------------------------------------------------------------------

::: callout-important
<!--# update with team instructions-->

Make sure to do the following as you complete the assignment:

-   Write all code and narrative in your Quarto file. We should be able to read all your code in the rendered PDF.
-   Write all narrative in complete sentences.
-   Use informative axis titles and labels on all graphs.
-   Implement version control in your reproducible workflow. This means throughout the assignment, you should periodically **render** your Quarto document to produce the updated PDF, **commit** the changes in the Git pane, and **push** the updated files to GitHub. You should aim to push changes to GitHub at least three times as you work on the assignment.
:::

**Goal**: The goal of this analysis is to use linear regression to understand variability in overall quality rating based on the aftertaste rating in coffee.

### Exercise 1

Visualize the relationship between the aftertaste grade and total cup points. Write two observations from the plot.

### Exercise 2

Fit the linear model using the aftertaste grade to understand variability in the total cup points. Neatly display the model using three digits and include the **98%** confidence interval for the model coefficients in the output.

### Exercise 3

a.  Interpret the slope in the context of the data.
b.  Assume you are a coffee drinker. Would you drink a coffee represented by the intercept? Why or why not?

### Exercise 4

a.  Write code to "manually" compute the residuals. Save the residuals but do not print them out.
b.  Write code to compute the regression standard error, $\hat{\sigma}_\epsilon$ using the residuals from part (a).
c.  State the definition of the regression standard error in the context of the data.

::: callout-tip
You can check your answer to part (b) by using the code below to obtain $\hat{\sigma}_\epsilon$<center>

`glance(model_name)$sigma`

</center>
:::

### Exercise 5

Do the data provide evidence of a statistically significant linear relationship between aftertaste grade and total cup points? Conduct a hypothesis test using mathematical models to answer this question.

a.  State the null and alternative hypotheses in words and in mathematical notation.
b.  What is the test statistic? State what the test statistic means in the context of this problem.
c.  What distribution was used to calculate the p-value? Be specific.
d.  State the conclusion in the context of the data using a threshold of $\alpha = 0.02$ to make your decision.

### Exercise 6

a.  What is the critical value used to calculate the confidence interval displayed in Exercise 2? Show the code and output used to get your response.

b.  Is the confidence interval consistent with the conclusions from the hypothesis test? Briefly explain why or why not.

### Exercise 7

-   Calculate the 98% confidence interval for the mean total cup points grade for coffees with aftertaste grade of 8.25. Interpret this value in the context of the data.

-   One coffee produced by the Ethiopia Commodity Exchange has an aftertaste of 8.25. Calculate the the 98% prediction interval for the total cup points grade for this coffee. Interpret this value in the context of the data.

-   How do the predicted values compare? How do the intervals compare? If there are differences in the predictions and/or intervals, briefly explain why.

### Exercise 8

We have conducted the inference in the previous exercises making some inherent assumptions about the data. Therefore, we will check some model conditions to assess whether the assumptions and hold and the reliability of the inferential results. To do so, we will create a data frame called `coffee_aug` that includes the residuals and predicted values from the model. Input the name of your model in the code below.

```{r}
#|eval: false
coffee_aug <- augment(_____)
```

Make a scatterplot of the residuals (`.resid`) vs. fitted values (`.fitted`). Use [`geom_hline()`](https://ggplot2.tidyverse.org/reference/geom_abline.html) to add a horizontal dotted line at $residuals = 0$.

::: callout-note
The **linearity condition** is satisfied if there is random scatter of the residuals (no distinguishable pattern or structure) in the plot of residuals vs. fitted values.\
\
The **constant variance** condition is satisfied if the vertical spread of the residuals is relatively across the plot.
:::

a.  Is the linearity condition satisfied? Briefly explain why or why not.

b.  Is the constant variance condition satisfied? Briefly explain why or why not.

### Exercise 9

::: callout-note
The **normality** condition is satisfied if the distribution of the residuals is approximately normal. This condition can be relaxed if the sample size is sufficiently large $(n > 30)$.
:::

Make a histogram or density plot of the residuals (`.resid`). Is the normality condition satisfied? Briefly explain why or why not.

If not, can it be relaxed for this data set?

### Exercise 10

::: callout-note
The **independence** condition means that knowing one residual will not provide information about another. We often check this by assessing whether the observations are independent based on what we know about the subject matter and how the data were collected.

This condition is sometimes difficult to fully assess, so we just want to consider whether it is reasonably satisfied based on the information and data available.
:::

Is the independence condition satisfied for these data? Briefly explain why or why not.

# Submission

# <!--# update with team instructions-->

::: callout-warning
Before you wrap up the assignment, make sure all documents are updated on your GitHub repo. We will be checking these to make sure you have been practicing how to commit and push changes.

Remember -- you must turn in a PDF file to the Gradescope page before the submission deadline for full credit.
:::

To submit your assignment:

-   Access Gradescope through the [STA 210 Canvas site](https://canvas.duke.edu/courses/47067).

-   Click on the assignment, and you’ll be prompted to submit it.

-   Mark the pages associated with each exercise. All of the pages of your lab should be associated with at least one question (i.e., should be “checked”).

-   Select the first page of your .PDF submission to be associated with the *“Workflow & formatting”* section.

# Grading (50 points)

<br>

| Component             | Points |
|:----------------------|:-------|
| Ex 1                  |        |
| Ex 2                  |        |
| Ex 3                  |        |
| Ex 4                  |        |
| Ex 5                  |        |
| Ex 6                  |        |
| Ex 7                  |        |
| Ex 8                  |        |
| Ex 9                  |        |
| Ex 10                 |        |
| Workflow & formatting |        |
